Machine Learning guided Design\\ of Ribosome Binding Sites

### 2021-06-15 paper/main.tex
Final paper update.

% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
The ribosome binding site (RBS), as one of the transcription and translation control elements, allows fine control of gene expression.
% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
We consider the problem of designing RBSs at the nucleotide level to optimise gene expression with a fixed budget. 
To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
The algorithms were able to identify RBS sequences rules conducive to high protein expression.
Notably, by testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.

@@

@@

In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
% In short, this algorithm is a stochastic method of probing of the experimental space. 
% It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
To represent RBSs and capture the similarities between them, we chose to use the
Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\

>>>>

Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
The algorithms were able to identify RBS sequences rules conducive to high protein expression.
Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.

@@

The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
In short, this algorithm is a stochastic method of probing of the experimental space. 
It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\

@@

In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
To represent RBSs and capture the similarities between them, we chose to use the
Weighted Degree Kernel with Shift method, which fits well into our prediction method.
In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\

### 2021-06-15 paper/methods.tex
Final paper update.

Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequences sequentially (i.e. one-by-one) and then wait for the label to be tested and used to improve the model.
Instead, we can recommend RBS sequences in a batch of size $m$. 
One naive approach is to 
%train GP on sequences with known TIR labels and
recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}(A) ($m = 2$).
However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
Since GPR assumes similar sequences would have similar labels (e.g. by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Point_Top_2_UCB.pdf}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_1st_Data_Point.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_2nd_Data_Point.pdf}
    \end{subfigure}
    \caption{\emph{Batch Recommendation Illustration.} We use the batch size of 2, with 5 initial observations. The design space is 24 uniformly distributed points in the range [-2,10], i.e. {-2, -1.5, -1, ..., 9.5, 10}.
    The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions. 
    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that.}
    \label{fig:batch rec}
\end{figure}

A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), not on the labels of those observed points. 
% We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
% As we can see, the predicted confidence interval is the same for the two plots. 
One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
The strategy here is to recommend sequences sequentially by updating the UCB score with the predicted standard deviation of the previously recommended points.\\
% Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
% The plot shows the predicted mean and standard deviation after observing 4 data points. 
As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e. observation), and update the predicted standard deviation and then we finally update the UCB scores. 
The second data point is then recommended based on the new UCB scores.   
Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, the algorithm recommends $x = 8$, which is in another local maximum design area. 
In this way, the batch recommendation can potentially cover more local maximum areas than the sequential design. 
In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 
In our design, we have set the batch size as 90, to fit the experimental batch.
We set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.

>>>>

Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequences sequentially (i.e. one-by-one) and then wait for the label to be tested and used to improve the model.
Instead, we can recommend RBS sequences in a batch of size $m$. 
One naive approach is to 
%train GP on sequences with known TIR labels and
recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}A ($m = 2$).
However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
Since GPR assumes similar sequences would have similar labels (e.g. by knowing $x=2$ we can gain information about $x=2.5$ as well), we prefer to not to waste time and money on labelling sequences with high similarities in the same batch.
To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Point_Top_2_UCB.pdf}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_1st_Data_Point.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_2nd_Data_Point.pdf}
    \end{subfigure}
    \caption{\emph{Batch Recommendation Illustration.} We use the batch size of 2, with 5 initial observations. The design space is 24 uniformly distributed points in the range [-2,10], i.e. {-2, -1.5, -1, ..., 9.5, 10}.
    The confidence intervals are shown with predicted mean of $\pm$ 1.96 standard deviation.
    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions. 
    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that updated interval.}
    \label{fig:batch rec}
\end{figure}

A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), not on the labels of those observed points. 
% We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
% As we can see, the predicted confidence interval is the same for the two plots. 
One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
The strategy here is to recommend sequences sequentially by updating the UCB score with the predicted standard deviation of the previously recommended points.\\
% Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
% The plot shows the predicted mean and standard deviation after observing 4 data points. 
As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e. observation), and update the predicted standard deviation and then we finally update the UCB scores. 
The second data point is then recommended based on the new UCB scores.   
Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, the algorithm recommends $x = 8$, which is in another local maximum design area. 
In this way, the batch recommendation can potentially cover more local maximum areas than the sequential design. 
In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 
In our design, we have set the batch size to 90, to fit the experimental batch.
Finally, we set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.
